<small> Supplementary materials for the main submission entitled - <br><br> A new metric for automatic discovery of periodic patterns in time series

### 2021-06-01 paper/main.Rmd
updated all simulation files

Exploratory data analysis, as coined by John W. Tukey (Tukey 1965) involves many iterations of finding structures and patterns that allow the data to be informative. With temporal data available at finer scales, exploring time series data can become overwhelming with so many possible cyclic temporal granularities [@Gupta2020-vo], which are temporal deconstructions that represent cyclic repetitions in time, e.g. hour-of-day, day-of-month, or any public holidays. These granularities form ordered (for eg. day-of-week where Tue is always followed by Wed, which again is followed by Thu and so on) or unordered categorical variables. Therefore, exploring univariate time series data amounts to exploring the distribution of the measured variable across different categories of the cyclic granularities. \noindent Take the example of the  electricity smart meter data used in @wang2020calendar for four households in Melbourne, Australia. Figure \ref{fig:onegran} shows the distribution of energy usage of one household (id 2) across cyclic granularity a) hour-of-day and b) month-of-year. Potentially many such displays could be drawn across day-of-week, day-of-month, weekday/weekend, or any other chosen cyclic granularities of interest. However, all of them would not be interesting to discern important patterns in the energy usage. Only those displays, which have "significant" distributional differences between categories of the cyclic granularity would be informative.

@@

```{r onegran, fig.cap="Boxplots showing the distribution of one household across one cyclic granularity at a time - (a) hour-of-day and (b) and month-of-year. The daily and annual periodic behavior of energy is apparent in (a) and (b) respectively, with daily peaks occuring in morning and evening hours when members in the house are present and active, more volatility (usage of air conditioner) in summer months (jan, feb) due to air conditioners and more consistent behavior in winter and autumn months (mar, apr, may, june). "}
```

<!-- comparing one and two grans for one households -->
\noindent Exploring the distribution of the measured variable across two cyclic granularities tend to provide more detailed information on its structure. Figure \ref{fig:id2}(a) shows that energy consumption is higher during the morning hours (5-8) when members in the household wake up and then again in the evening hours (17-20) possibly when members get back from work with maximum variation (large inter-quartile range) in behavior in the afternoon hours (12-16). Figure \ref{fig:id2}(b) shows the distribution of energy consumption across months January to June. The median and quartile deviation of energy usage in Jan and Feb are generally on a much higher side, possibly due to the usage of air conditioners (Jan, Feb are peak summer in Australia), however for other months (Mar-Jun, autumn and winter), the smaller median and quartile deviation indicate a more consistent behavior. It might also imply that this household do not use as much heater as compared to air conditioner. A lot of households in Victoria use gas heating and hence the usage of heaters might not be reflected here. Figure \ref{fig:id2}(a) slice it down further by showing the usage distribution across hour-of-day conditional on month-of-year. It shows the hourly usage over a day do not remain same across months. Unlike other months, the 75th and 90th percentile for all hours of the day in January are high, pretty close and are not characterized by a morning and evening peak. The household in Figure \ref{fig:id2}(b) has 90th percentile consumption higher in summer months relative to autumn or winter, but the 75th and 90th percentile are far apart in all months, implying that the second household resorts to air conditioning much less regularly than the first one. The differences seem to be more prominent across month-of-year (facets) than hour-of-day (x-axis) for this households, whereas they are prominent for both cyclic granularities for the first household. It could be immensely useful to make the transition from all potential displays to the ones that are informative across atleast one cyclic granularity.

@@

# Ranking and selecting significant harmonies {#sec:rank-wpd}

A harmony is referred to as "significant" if there is a significant distributional difference of the measured variable between different categories of the harmony. In this section, a selection criterion to choose significant harmonies is provided, thereby eliminating all harmonies that exhibit complete randomness in the measured variable. The distance measure $wpd$ is used as a test statistic to test the null hypothesis that any given harmony is not significant. We select only those harmonies for which the test fails. The significant harmonies are then ranked basis how well they capture variation in the measured variable.

@@

## Choosing a threshold and selection criterion

@@

Figure \ref{fig:raw} shows that both the location and scale of the distributions change across panels. This is not desirable under $D_{null}$ as it would mean comparisons of $wpd$ values is not appropriate across different $nx$ and $nfacet$. \ref{tab:glm-tab} gives the summary of the generalised linear model to capture the relationship between $wpd_{raw}$ and number of comparisons. Thus, the model considered is $wpd_{l,s} = 23.69448+-1.02357*log(nx * nfacet)) + e$. The intercepts are similar independent of the starting distribution (Please refer to supplementary).
Figure \ref{fig:dist-new-same-scale-link} shows the distribution of $wpd_{perm}$ and $wpd_{glm-scaled}$ in the same scale to show that a combination approach could be used for higher values of levels to alleviate the computational time of permutation approach.

@@

### Design

Simulations are conducted to understand the efficacy of the test. Observations are generated from a N(0,1) distribution for each combination of $nx$ and $nfacet$ from the following sets: $nx = \{3, 7, 14\}$ and $nfacet = \{2, 9, 10\}$. The panel $(3, 2), (7,9), (14, 10)$ are considered to have design $D_{null}$. The panels $(7, 2), (14, 9)$ have design of the form $D_{var_f}$. $(14, 2), (3, 10)$ have design of the form $D_{var_x}$ and the rest are under $D_{var_{null}}$. We generate only one data set for which all these designs were simulated and consider this as the original data set. We generate $200$ repetitions of this experiment with different seeds and compute the proportion of times a panel is rejected when it is under $D_{null}$. We also compute the proportion of times a panel is rejected when it actually belongs to a non-null design. The first proportion is desired to be as small as possible and a higher value of the later is expected. Also, these would constitute to be the estimated size and power of the test. The results are included in details in the Supplementary paper.

>>>>

Exploratory data analysis, as coined by John W. Tukey [@Tukey1977-jx] involves many iterations of finding structures and patterns that allow the data to be informative. With temporal data available at finer scales, exploring time series data can become overwhelming with so many possible cyclic temporal granularities [@Gupta2020-vo], which are temporal deconstructions that represent cyclic repetitions in time, e.g. hour-of-day, day-of-month, or any public holidays. These granularities form ordered (for eg. day-of-week where Tue is always followed by Wed, which again is followed by Thu and so on) or unordered categorical variables. Therefore, exploring univariate time series data amounts to exploring the distribution of the measured variable across different categories of the cyclic granularities. \noindent Take the example of the  electricity smart meter data used in @wang2020calendar for four households in Melbourne, Australia. Figure \ref{fig:onegran} shows the distribution of energy usage of one household (id 2) across cyclic granularity a) hour-of-day and b) month-of-year. Potentially many such displays could be drawn across day-of-week, day-of-month, weekday/weekend, or any other chosen cyclic granularities of interest. However, all of them would not be interesting to discern important patterns in the energy usage. Only those displays, which have "significant" distributional differences between categories of the cyclic granularity would be informative.

@@

```{r onegran, fig.cap="Boxplots showing the distribution of one household across one cyclic granularity at a time - (a) hour-of-day and (b) and month-of-year. The daily and annual periodic behavior of energy is apparent in (a) and (b) respectively, with daily peaks occuring in morning and evening hours when members in the house are present and active, more volatility (usage of air conditioner) in summer months (Jan, Feb) due to air conditioners and more consistent behavior in winter and autumn months (Mar, Apr, May, June)."}
```

<!-- comparing one and two grans for one households -->
\noindent Exploring the distribution of the measured variable across two cyclic granularities tend to provide more detailed information on its structure. Figure \ref{fig:onegran}(a) shows that energy consumption is higher during the morning hours (5-8) when members in the household wake up and then again in the evening hours (17-20) possibly when members get back from work with maximum variation (large inter-quartile range) in behavior in the afternoon hours (12-16). Figure \ref{fig:onegran}(b) shows the distribution of energy consumption across months January to June. The median and quartile deviation of energy usage in Jan and Feb are generally on a much higher side, possibly due to the usage of air conditioners (Jan, Feb are peak summer in Australia), however for other months (Mar-Jun, autumn and winter), the smaller median and quartile deviation indicate a more consistent behavior. It might also imply that this household do not use as much heater as compared to air conditioner. A lot of households in Victoria use gas heating and hence the usage of heaters might not be reflected here. Figure \ref{fig:id2}(a) slice it down further by showing the usage distribution across hour-of-day conditional on month-of-year across two households (id 2 and 4). It shows the hourly usage over a day do not remain same across months. Unlike other months, the 75th and 90th percentile for all hours of the day in January are high, pretty close and are not characterized by a morning and evening peak. The household in Figure \ref{fig:id2}(b) has 90th percentile consumption higher in summer months relative to autumn or winter, but the 75th and 90th percentile are far apart in all months, implying that the second household resorts to air conditioning much less regularly than the first one. The differences seem to be more prominent across month-of-year (facets) than hour-of-day (x-axis) for this households, whereas they are prominent for both cyclic granularities for the first household. It could be immensely useful to make the transition from all potential displays to the ones that are informative across atleast one cyclic granularity.

@@

# Ranking and selection of cyclic granularities {#sec:rank-wpd}

A cyclic granularity is referred to as "significant" if there is a significant distributional difference of the measured variable between different categories of the harmony. In this section, a selection criterion to choose significant harmonies is provided, thereby eliminating all harmonies that exhibit complete randomness in the measured variable. The distance measure $wpd$ is used as a test statistic to test the null hypothesis that any given harmony is not significant. We select only those harmonies for which the test fails. The significant harmonies are then ranked basis how well they capture variation in the measured variable.

@@

## Selection

@@

Figure \ref{fig:raw} shows that both the location and scale of the distributions change across panels. This is not desirable under $D_{null}$ as it would mean comparisons of $wpd$ values is not appropriate across different $nx$ and $nfacet$. \ref{tab:glm-tab} gives the summary of the generalized linear model to capture the relationship between $wpd_{raw}$ and number of comparisons. Thus, the model considered is $wpd_{l,s} = 23.69448+-1.02357*log(nx * nfacet)) + e$. The intercepts are similar independent of the starting distribution (Please refer to supplementary).
Figure \ref{fig:dist-new-same-scale-link} shows the distribution of $wpd_{perm}$ and $wpd_{glm-scaled}$ in the same scale to show that a combination approach could be used for higher values of levels to alleviate the computational time of permutation approach.

@@

Observations are generated from a N(0,1) distribution for each combination of $nx$ and $nfacet$ from the following sets: $nx = \{3, 7, 14\}$ and $nfacet = \{2, 9, 10\}$. This would result in 9 panels, viz, $(3, 2), (3, 9), (3, 10), \dots, (14,9), (14, 10)$. Few experiments were conducted. In the first scenario, data for all panels are simulated using the null design  $D_{null}$. In other scenarios, data simulated from the panel $(14, 2)$ and $(3, 10)$ are under $D_{vary_{all}}$. Moreover, $\omega = \{0.5, 2, 5\}$ are considered to examine if the proposed test is able to capture subtle differences and non-subtle differences when we shift from the null design. In the last scenario, we consider the panel $(3, 2), (7,9), (14, 10)$ to be under $D_{null}$, the panels $(7, 2), (14, 9)$ to be under $D_{var_f}$. $(14, 2), (3, 10)$ under $D_{var_x}$ and the rest under $D_{var_{null}}$. This is done to check if the consequent ranking procedure leads to designs like $D_{vary_{all}}$ to be chosen first followed by $D_{vary_{all}}$. We generate only one data set each for which these scenarios were simulated and consider this as the original data set. We generate $1000$ repetitions of this experiment with different seeds.

The proportion of times a panel is rejected when it is under $D_{null}$ is computed with $wpd_{threshold99}$ and it is around $0.1$. This is obvious, since if the level of significance for each test is less than $0.01$ (as a result of choosing $99^{th}$ percentile as the threshold) and we have $9$ tests, the size for the entire test should be close to $0.01 *9 \approx 0.1$. We also compute the proportion of times a panel is rejected when it actually belongs to a non-null design. The first proportion is desired to be as small as possible and a higher value of the later is expected. Also, these would constitute to be the estimated size and power of the test. It is found that as we increase from low to high changes from the null distribution, the power increased. The results and graphics are included in details in the Supplementary paper.

```{r, eval = FALSE}
all_data <- read_rds("simulations/supplementary/test-hpc/data-agg/all_data.rds")

# computes size for 99 percentile threshold
data_summary <- all_data %>%
  mutate(sig = if_else(significance!=99, 0, 99)) %>% 
  group_by(seed_id) %>% 
  summarize(sumsig = sum(sig)) %>% 
  count(sumsig==0) %>% 
  mutate(p_value = n/sum(n)) %>% 
  slice(1)

# plots
ggplot() + 
  geom_histogram(data = all_data, aes(x = wpd))  + 
  geom_vline(data = data_summary, aes(xintercept = p_value),colour = "red") +
  facet_grid(x_levels~facet_levels)

### 2021-06-01 paper/paper-supplementary.Rmd
updated all simulation files

>>>>

```{r hist-qq-new, fig.cap = "In panel a, the histogram of $wpd_{glm-scaled}$ is plotted. In part b, the QQ plot is shown with the theoretical quantiles on the x-axis and $wpd_{glm-scaled}$ quantiles on the y-axis. The distribution looks symmetric and looks like normal except in the tails.", eval = FALSE}

```

```{r quadratic, fig.cap = "$wpd$ is plotted against $nx*nfacet$ (the maximum number of pairwise comparisons) and the blue line represents the median of the multiple values for each $nx*nfacet$. The median increases abruptly for lower values of $nx*nfacet$ and slowly for higher $nx*nfacet$. Thus, the measure will have higher values for higher levels in $nx$ or $nfacet$.", eval = FALSE}
```

```{r norm, fig.cap = "Distribution of $wpd_{perm}$ is plotted across different $nx$ and $nfacet$ categories. Both shape and scale of the distributions are now similar for different panels under the null design."}
```

```{r wpd-glm-dist, fig.cap = " The distribution of $wpd_{glm}$ is plotted. The distributions are more similar across higher nx and nfacet and dissimilar for fewer nc and nfacets."}

```

$$y_l = a+b*log(z_l) + e_l , \quad l = (1, 2, \dots, L)$$,

\noindent where $y_l = median_s(wpd_{l, s})$, $wpd_{l, s}$ denotes the value of $wpd_{raw}$ obtained for the $l^{th}$ panel and $s^{th}$ simulation, $z_l$ is the number of groups $(nx*nfacet)$ in the $l^{th}$ panel and $e_l$ are idiosyncratic errors. Let $E(y) = \mu$ and $a + b*log(z) = g(\mu)$ where $g$ is the link function. Then $g(\mu)=  1/\mu$ and $\hat \mu  = 1/(\hat a + \hat b log(z))$. The residuals from this model $(y-\hat y) = (y-1/(\hat a + \hat b log(z)))$ would be expected to have no dependency on $z$. Thus, $wpd_{glm}$ is chosen as the residuals from this model and is defined as:
$wpd_{glm} = wpd_{raw} - 1/(\hat a + \hat b*log(nx*nfacet))$.
